Cooperative Kernels: GPU Multitasking for Blocking Algorithms

### 2017-02-27 write-ups/FSE/paper.tex
added new results and descriptions around them

\begin{figure*}
\includegraphics[width=.67\columnwidth]{iris_octree_NA.pdf}
\includegraphics[width=.67\columnwidth]{hd520_mis_ecology.pdf}
\includegraphics[width=.67\columnwidth]{hd5500_bfs_rmat22.pdf}
%\vspace{-2mm}
\caption{Example gather time and non-cooperative timing results \TSComment{need to get updated numbers}}\label{fig:fine-grained-timing}
\vspace{-3mm}
\end{figure*}

\begin{table}
\normalsize
\centering
\begin{tabular}{ l l | l l | l l }
\multicolumn{2}{c|}{overall} & \multicolumn{2}{c|}{barrier} & \multicolumn{2}{c}{wk.steal.} \\
mean & max & mean & max & mean & max \\
\hline
$1.09$ & $1.23^{\ddagger}$ & $1.09$ & $1.21^{\diamond}$ & $1.11$ & $1.23^{\ddagger}$ \\
\end{tabular}\\
{\small
 $^{\ddagger}$octree, $^{\diamond}$sssp usa\_ny
}
%\vspace{-2mm}
\caption{Cooperative kernel slowdown w/o multitasking \TSComment{need to get updated numbers}}
\label{tab:overhead}
\vspace{-8mm}
\end{table}

\myparagraph{Results}
\mytab\ref{tab:overhead} shows the geometric mean and
maximum slowdown across all applications and inputs, with averages and
maxima computed over 10 runs per benchmark. For the maximum slowdowns,
we indicate which application and input was responsible. The slowdown is
below 1.5 even in the worst case, and closer to 1 on average. We consider
these results encouraging, especially since the performance of our
prototype could clearly be improved upon in a native implementation.

@@

Our experiments span 13 pairs of cooperative kernels and inputs, 3
classes of non-cooperative kernel workloads, 4 quantities of
workgroups claimed for the non-cooperative kernel and 2 variations of
resizing barriers for graph algorithms, leading to 288 configurations.
We run each configuration 10 times, in order to report averaged
performance numbers. For each run, we record the execution time of the
cooperative kernel. For each scheduling of the non-cooperative kernel
during the run, we also record the \emph{gather time} needed by the
scheduler to collect enough workgroups to launch the non-cooperative
kernel, and the non-cooperative kernel execution time.
%
%We sanity-check
%whether all kernels perform their computations correctly by comparing
%their result with reference results.

@@

The middle and right graphs show results for graph algorithms.  These
algorithms use barriers, and we experimented with the regular and
query barrier implementations described in
\mysec\ref{sec:resizingbarrier}.  The execution times for the
non-cooperative task are averaged across all runs, including with both
types of barrier.  We show separately the average gather time
associated with each type of barrier.  The graphs show a similar trend
to the left-most graph: as the number of non-cooperative workgroups
grows, the execution time decreases and the gather time
increases.%\footnote{The right figure has only 3 points per line:
  %recall that for bfs there are no runs with ``all-but-one''
  %non-cooperative workgroups.}
%
the gather time is higher on the right figure: the rmat22 input graph
is rather wide than deep, so the graph algorithm reaches resizing
barriers less often than for the ecology input of the middle figure
for instance. The scheduler thus has fewer opportunities to collect
workgroups and gather time increases. Nonetheless, scheduling
responsiveness can benefit from the query barrier: when used, this
barrier lets the scheduler collect all needed workgroups as soon as
they hit a resizing barrier. As we can see, the gather time of the
query barrier is almost stable w.r.t.\ the number of workgroups that
needs to be collected.

\begin{figure}
\includegraphics[width=\columnwidth]{heavy.pdf}
%\vspace{-2mm}
\caption{Performance impact of multitasking cooperative and non-cooperative workloads, and the period with which non-cooperative kernels execute \TSComment{need to get updated numbers}}\label{fig:performance}
\vspace{-5mm}
\end{figure}

\myparagraph{Performance} \myfiglong\ref{fig:performance} reports the
overhead brought by the scheduling of non-cooperative kernels over the
cooperative kernel execution time.  This is the slowdown associated
with running the cooperative kernel in the presence of multitasking,
vs.\ running the cooperative kernel in isolation (geometric mean over
all applications and inputs).  We also show the period at which
non-cooperative kernels can be scheduled (arithmetic mean over all
applications and inputs).
We show results for the three workloads listed in
\mysec~\ref{sec:sizingnoncoop}. The two gray horizontal lines in the
period graph correspond to the period goals of the workloads: the
higher (resp. lower) line corresponds to a period of 70ms (resp. 40ms)
for the light (resp. medium and heavy) workload.

Co-scheduling non-cooperative kernels that request a single workgroup
leads to almost no overhead, but the period is far too high to meet
the needs of any of our three workloads; e.g.\ a heavy workload
averages a period of 939 ms. As more workgroups are dedicated to
non-cooperative kernels, they execute quickly enough to be scheduled
at the expected period. For the light and medium workloads, a quarter
of the workgroups executing the non-cooperative kernel are able to
meet their goal period (70 and 40 ms resp.). However, a quarter of the
workgroups are not sufficient to meet the goal for the heavy workload
(giving a mean period of 88 ms). If half of the workgroups are
allocated to the non-cooperative kernel, the heavy workload averages
$\sim$10\% over its goal period (mean of 44 ms). If all-but-one are
allocated, the heavy workload reaches its goal period.
%
Yet, as expected, allocating more non-cooperative workgroups increases
the execution time of the cooperative kernel.

Still, heavy workloads meet their period by allocating all-but-one
non-cooperative workgroups, incurring a slow down of less than
1.4$\times$ on average. Light and medium workloads meet their period
with only a negligible overhead of the cooperative kernel (less than a
1.03$\times$ slowdown on average).

@@

\begin{table}[t]
\normalsize
\centering
\begin{tabular}{ l r r}
g. workload &  kernel-level  & cooperative (resources) \\
\hline
light & 1.04 & 1.XX($N/4$)\\
medium & 1.08 & 1.XX($N/4$)\\
heavy & 1.33 & 1.XX($N-1$)\\
\end{tabular} \\
%\vspace{.2cm}

\caption{Overhead of kernel level preemption vs cooperative kernels for three
graphics workloads \TSComment{need to get updated numbers}}
\label{tab:preemption}
\vspace{-8mm}
\end{table}

@@

Let us assume that a graphics workload is required to be scheduled
with period $P$ and duration $D$, and that a long-running compute
kernel requires time $C$ to execute without interruption.  If we
assume the cost of preemption is negligible (e.g.\ Nvidia have
reported preemption times of 0.1 ms for Pascal~\cite{TODO}), then the overhead
associated with switching between compute and graphics every $P$ time
steps is $P/(P-D)$.

We compare this task-level preemption overhead
model with our experimental results per graphics workload in
\mytab{\ref{tab:preemption}}.
%
Based on the above assumptions, our approach provides lower overhead for low and medium
graphics workloads. Recall for the heavy graphics workload, we are
within $10\%$ of the required deadline ($40$ ms) by using half the
workgroups with an overhead of $1.XX$, which is less than the
kernel-level preemption overhead of $1.33$. \TSComment{need to get
  updated numbers} \ADComment{I don't quite understand the wording; sounds a bit like we still win with the heavy workload.}

The low overhead of cooperative kernels for low and medium graphics
workloads is due to sub-linear scalability of the irregular algorithms that they implement: some resources can
be taken away during computation without proportionally reducing
execution time.  Our poor performance for the heavy workload is
because the graphics task requires all but one workgroup, which
cripples the cooperative kernel enough that $\offerfork$ calls are not
issued as frequently.
%
These results suggest that a hybrid preemption scheme may work the
best for multi-tasking and graphics. That is, the cooperative approach
works well for light and medium tasks, but heavy graphics tasks
benefit from the coarser grained, kernel-level preemption strategy.

>>>>

\begin{figure*}
\includegraphics[width=.67\columnwidth]{iris_octree_NA.pdf}
\includegraphics[width=.67\columnwidth]{iris_bfs_usa.pdf}
\includegraphics[width=.67\columnwidth]{iris_color_G3_circuit.pdf}
%\vspace{-2mm}
\caption{Example gather time and non-cooperative timing results}\label{fig:fine-grained-timing}
\vspace{-3mm}
\end{figure*}

\begin{table}
\normalsize
\centering
\begin{tabular}{ l l | l l | l l }
\multicolumn{2}{c|}{overall} & \multicolumn{2}{c|}{barrier} & \multicolumn{2}{c}{wk.steal.} \\
mean & max & mean & max & mean & max \\
\hline
$1.07$ & $1.23^{\ddagger}$ & $1.06$ & $1.20^{\diamond}$ & $1.12$ & $1.23^{\ddagger}$ \\
\end{tabular}\\
{\small
 $^{\ddagger}$octree, $^{\diamond}$color G3\_circuit
}
%\vspace{-2mm}
\caption{Cooperative kernel slowdown w/o multitasking}
\label{tab:overhead}
\vspace{-8mm}
\end{table}

\myparagraph{Results}
\mytab\ref{tab:overhead} shows the geometric mean and
maximum slowdown across all applications and inputs, with averages and
maxima computed over 10 runs per benchmark. For the maximum slowdowns,
we indicate which application and input was responsible. The slowdown is
below 1.25 even in the worst case, and closer to 1 on average. We consider
these results encouraging, especially since the performance of our
prototype could clearly be improved upon in a native implementation.

@@

Our experiments span 11 pairs of cooperative kernels and inputs, 3
classes of non-cooperative kernel workloads, 4 quantities of
workgroups claimed for the non-cooperative kernel and 2 variations of
resizing barriers for graph algorithms, leading to 240 configurations.
We run each configuration 10 times, in order to report averaged
performance numbers. For each run, we record the execution time of the
cooperative kernel. For each scheduling of the non-cooperative kernel
during the run, we also record the \emph{gather time} needed by the
scheduler to collect enough workgroups to launch the non-cooperative
kernel, and the non-cooperative kernel execution time.
%
%We sanity-check
%whether all kernels perform their computations correctly by comparing
%their result with reference results.

@@

The middle and right graphs show results for graph algorithms.  These
algorithms use barriers, and we experimented with the regular and
query barrier implementations described in
\mysec\ref{sec:resizingbarrier}.  The execution times for the
non-cooperative task are averaged across all runs, including with both
types of barrier.  We show separately the average gather time
associated with each type of barrier.  The graphs show a similar trend
to the left-most graph: as the number of non-cooperative workgroups
grows, the execution time decreases and the gather time
increases.%\footnote{The right figure has only 3 points per line:
  %recall that for bfs there are no runs with ``all-but-one''
  %non-cooperative workgroups.}
%
the gather time is higher on the right figure: the G3 circuit input
graph is rather wide than deep, so the graph algorithm reaches
resizing barriers less often than for the usa road input of the middle
figure for instance. The scheduler thus has fewer opportunities to
collect workgroups and gather time increases. Nonetheless, scheduling
responsiveness can benefit from the query barrier: when used, this
barrier lets the scheduler collect all needed workgroups as soon as
they hit a resizing barrier. As we can see, the gather time of the
query barrier is almost stable w.r.t.\ the number of workgroups that
needs to be collected.

\begin{figure}
\includegraphics[width=\columnwidth]{heavy.pdf}
%\vspace{-2mm}
\caption{Performance impact of multitasking cooperative and non-cooperative workloads, and the period with which non-cooperative kernels execute}\label{fig:performance}
\vspace{-5mm}
\end{figure}

\myparagraph{Performance} \myfiglong\ref{fig:performance} reports the
overhead brought by the scheduling of non-cooperative kernels over the
cooperative kernel execution time.  This is the slowdown associated
with running the cooperative kernel in the presence of multitasking,
vs.\ running the cooperative kernel in isolation (median over all
applications and inputs).  We also show the period at which
non-cooperative kernels can be scheduled (median over all applications
and inputs). We use the median as our data included some drastic
outliers; these outliers appear to happen with applications/inputs in
which the resizing barrier are not called very frequently and the
graphics task asks for half or more resources. For example, a medium
graphics workload for bfs on the rmat input has over an 8$\times$
overhead, while the same application on the usa input has an overhead
of 1.11$\times$. As \myfiglong\ref{fig:performance} shows, most of our
applications/inputs are much better behaved than this. Future work
includes examining these problem applications/inputs in more detail,
possibly to insert more calls to $\offerfork$ and $\offerkill$.

We show results for the three workloads listed in
\mysec~\ref{sec:sizingnoncoop}. The two gray horizontal lines in the
period graph correspond to the period goals of the workloads: the
higher (resp. lower) line corresponds to a period of 70ms (resp. 40ms)
for the light (resp. medium and heavy) workload.

Co-scheduling non-cooperative kernels that request a single workgroup
leads to almost no overhead, but the period is far too high to meet
the needs of any of our three workloads; e.g.\ a heavy workload
averages a period of 939 ms. As more workgroups are dedicated to
non-cooperative kernels, they execute quickly enough to be scheduled
at the expected period. For the light and medium workloads, a quarter
of the workgroups executing the non-cooperative kernel are able to
meet their goal period (70 and 40 ms resp.). However, a quarter of the
workgroups are not sufficient to meet the goal for the heavy workload
(giving a median period of 104 ms). If half of the workgroups are
allocated to the non-cooperative kernel, the heavy workload achieves
its goal period (median of 40 ms). 
%
Yet, as expected, allocating more non-cooperative workgroups increases
the execution time of the cooperative kernel.

Still, heavy workloads meet their period by allocating half
non-cooperative workgroups, incurring a slow down of less than
1.5$\times$ (median). Light and medium workloads meet their period
with only a small overhead of the cooperative kernel; 
1.04$\times$ and 1.08$\times$ median slowdown respectively.

@@

\begin{table}[t]
\normalsize
\centering
\begin{tabular}{ l r r r}
g. workload &  kernel-level  & cooperative & resources \\
\hline
light & 1.04 & 1.04 & $N/4$\\
medium & 1.08 & 1.08 & $N/4$\\
heavy & 1.33 & 1.47 & $N/2$\\
\end{tabular} \\
%\vspace{.2cm}

\caption{Overhead of kernel level preemption vs cooperative kernels for three
graphics workloads}
\label{tab:preemption}
\vspace{-8mm}
\end{table}

@@

Let us assume that a graphics workload is required to be scheduled
with period $P$ and duration $D$, and that a long-running compute
kernel requires time $C$ to execute without interruption.  If we
assume the cost of preemption is negligible (e.g.\ Nvidia have
reported preemption times of 0.1 ms for Pascal~\cite{anandtech}), then
the overhead associated with switching between compute and graphics
every $P$ time steps is $P/(P-D)$.

We compare this task-level preemption overhead model with our
experimental results per graphics workload in
\mytab{\ref{tab:preemption}}. We report the overhead of the
configuration that allowed us to meet to deadline of the graphics
task.
%
Based on the above assumptions, our approach provides similar overhead
for low and medium graphics workloads; however, for the high workload,
our approach has a higher overhead.

Our poor performance for the heavy workload is because the graphics
task requires half of the workgroups, which cripples the cooperative
kernel enough that $\offerfork$ calls are not issued as frequently.
Future work may examine how to insert more resizing calls in these
applications to decrease this overhead.
%
These results suggest that a hybrid preemption scheme may work the
well for multi-tasking and graphics. That is, the cooperative approach
works well for light and medium tasks and requires no special hardware
or preemption state saving; on the other hand, heavy graphics tasks
benefit from the coarser grained, kernel-level preemption strategy.
